Revealing transient strain in geodetic data with Gaussian process regression

### 2017-11-10 ResponseToReviewers/ResponseToReviewers.tex

\textbf{Review of the paper : Revealing transient strain in geodetic data
with Gaussian process regression}

\textbf{Overview of the work: The paper focuses on a new non-parametric
Bayesian method to estimate transient strain rate from GNSS data. It
is based on Gaussian Process regression. The method also includes an
outlier removal processing step. This technique is then applied to the
detection of slow slip events in Cascadia. The validity of the
detected transient signals is accomplished via comparison with the
seismic tremor recorded along the Cascadia range.}

\textbf{The article is well-written and your algorithm is very interesting. I
think the use of Gaussian Process Regression is novel and open new
roads in GNSS data analysis. I believe that you spend a lot of efforts
in trying to justify most of the assumptions behind your model,
together with the effects of using different kind of prior. Perhaps,
that overshadows slightly your results. You could consider to
reorganize the article to make it more readable for geophysicists.}

I have taken several steps to make my revised draft more accessible to
someone who does not have a strong background in stochastic processes:
I rewritten the methods section to be clearer, avoid esoteric jargon,
contain more examples, and elaborate on potentially unfamiliar
concepts. I have also added more references to the methods section to
make it easier for the reader to find additional information. I
sincerely hope that my revised draft is more readable to
geophysicists.

@@

\textbf{Page 4: Stochastic model: It would seem that the assumption of no
covariance between the north and east components is not a very good
one since there is a tectonic framework in which the transients occur
and its unlikely this framework aligns along the Cardinal directions.
Maybe add a statement of impact of the neglect and possibly the idea
of reorienting the “axes” to align with the tectonic framework e.g.,
perpendicular to the subduction zone interface in this case?}

This is a good point that we neglected to mention in the previous
draft. We have added a paragraph to Section 2 that explains the
reasoning and implications for this assumption. The paragraph starts
with "It should be noted that we have ignored any covariances ...".

\textbf{Page 4/5 Equation (3): Is it worth discussing at this point whether
the estimates of u and n (eta) can be separated without explicit prior
knowledge (which we may not have).  For example, orbit modeling errors
on one satellite (e.g., due to an unknown yaw problem in the
spacecraft) wilt generate a spatially and temporally correlated error
in $d_{ij}$.  How does this (stochastically non-modeled) error not project
in the u estimates?}

If we have absolutely no prior knowledge about u or eta then it would
not be possible to distinguish transient displacements from noise. I
think this should go without saying.

In this paper, our "prior knowledge" is that inland stations are
unlikely to record SSE displacements, which allows us to isolate and
characterize the statistical properties of eta. In Section 3.1, we
only explored the temporal covariance of eta, but we could have also
explored the spatial covariance of eta and potentially identified
noise from orbit modeling errors. If we did find spatially correlated
noise in eta, then our estimates of u would not be invalidated;
rather, the estimates of u would just have larger uncertainties.
However, we did not look for spatially correlated noise in this paper,
and it is certainly possible that an unknown noise process could have
propagated into our estimates of u.

While the reviewer makes a valid point, I think it should go without
saying that the an inappropriate model for $u$ and $\eta$ would
invalidate our results for transient strain rates.

it is unclear how we should
update the text. In section 2 of the text, the statistical properties
of $u$ and $\eta$ are assumed known, although not yet specified.

It
should go without saying that an inappropriate model for $u$ or $\eta$
would invalidate our estimates of transient strain rates.

 We make it clear in the text that we do that

If there are unknown orbit modeling errors, for example, then that noise could conceivably be

Once eta is
characterized, we are able to characterize u.

 We then characterize the eta

If a spatially and temporally correlated noise exists and we have incorporated it into

### 2017-11-11 ResponseToReviewers/ResponseToReviewers.tex

\textbf{Page 4: Stochastic model: It would seem that the assumption of no
covariance between the north and east components is not a very good
one since there is a tectonic framework in which the transients occur
and its unlikely this framework aligns along the Cardinal directions.
Maybe add a statement of impact of the neglect and possibly the idea
of reorienting the “axes” to align with the tectonic framework e.g.,
perpendicular to the subduction zone interface in this case?}

This is a good point that we neglected to mention in the previous
draft. We have added a paragraph to Section 2 that explains the
reasoning and implications for this assumption. The paragraph starts
with "It should be noted that we have ignored any covariances ...".

\textbf{Page 4/5 Equation (3): Is it worth discussing at this point whether
the estimates of u and n (eta) can be separated without explicit prior
knowledge (which we may not have).  For example, orbit modeling errors
on one satellite (e.g., due to an unknown yaw problem in the
spacecraft) wilt generate a spatially and temporally correlated error
in $d_{ij}$.  How does this (stochastically non-modeled) error not project
in the u estimates?}

At this point in the text, it is assumed that the statistical
properties of u and eta are known, although not yet specified. So I do
not think this would be an appropriate place to discuss a scenario in
which we do not know u or eta. I also think that it should go with
saying that the validity of our results are contingent on having a
good model for u and eta.

In Section 3.1 and 3.2, we discuss how we choose a model for u and
eta. The key piece of "prior knowledge" that allows us to discern u
from eta is that u is negligible for inland stations. This allowed us
to islolate and constrain eta first, and then we constrained u. If we
allowed eta to be spatially correlated (which we did not), then we
would have conceivably been able to describe the spatially and
temporally correlated noise from, for example, orbit modeling errors.

@@

\textbf{Page 6: Equation 10: Maybe it is discussed later in the paper but it
is probably worthwhile stating at this point how the inverse is
performed with a block 0 in the matrix?}

The matrix being inverted in eq. 10 and 11 of the revised draft is
still invertible with the block 0, provided that the columns of
$G$ are linearly dependent. This is explained in the revised
draft.

@@

\textbf{Page 6: Equation 12: Maybe some additional explanation is
needed here. How are these partial derivatives formed.  My
understanding the u estimates at this point correspond one-to-one with
the positions and times of the GNSS position determinations.  Did I
miss something?}

The posterior displacements, $\hat{u}$, are actually a spatially and
temporally continuous stochastic process. We can evaluated $\hat{u}$
at any position and time that we may be interested in (not just the
position and times of the data). We are also able to compute the
spatial and temporal derivatives analytically. In Section 2 of the
revised draft, we explain how the derivatives are computed in greater
detail. We hope this clarifies any confusion.

@@

%% Editor Review
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\textbf{Editor:}\newline

\textbf{Editor’s Review of GJI MS 17.0593 ‘‘Revealing transient
strain...’’ by Hines and Hetland This is an novel algorithm applied to
a problem of great current interest. Based on the reviewer’s comments
and my own reading, I am recommending moderate revision. The authors’
response should include a description of their changes, and also a
version of the paper on which these can be seen (just as highlights in
the text).}

\textbf{Many of the reviewers’ comments address places where the paper
could be more clearly written, and my overall comment would be the
same. The current version perhaps spends a little too much time on the
background and not enough on explaining the method and its derivation.
As shown by the references below (these include some the authors
cite), and the Dermanis paper, which I have also attached, strain
estimation from scattered data is an area in which there has been a
lot more prior work than the authors mention. I am including these
references not because the authors need to cite many or even most of
them, but only to make the point that the problem of estimating
strain, and of looking for transients, is a familiar one: for many
readers, myself among them, much more familiar than the statistical
methods applied here. So a slightly fuller discussion of these methods
(perhaps a little more background and some reminders of, eg, what a
hyperparameter is) would be welcome to many readers; and increased
clarity of presentation usually means that the paper will be used
and cited more.}

I appreciate this comment, and I have extensively revised the paper in
order to make it more readable. To list some of the major changes:

  1) I properly define a Gaussian process in Section 2. I elaborate on
     its definition to give it a more tangible meaning, and I use
     Brownian motion as an example of a Gaussian process.

  2) I give more references, including section numbers for books, to
     make it easier for the reader to get additional information.

  3) In section 2 of the revised draft, I give an example of a prior
     for transient displacements. I include a figure showing what the
     covariance function looks like and what a corresponding
     realization looks like.

  4) I provide a clearer derivation of the posterior transient
     displacements (eq. 6 and 7 of the revised draft).

  5) In the revised draft, eq. 19 and 20 were added to help the reader
     understand how the transient strain rates are computed.

  6) I provide a clearer derivation of the signal-to-noise ratio.

  7) I moved details of the outlier detection method to the appendix
     because it is not essential to the text.

  8) I expanded the explanation of the REML method, which is used to
     constrain u and eta in Section 3.1 and 3.2.

  9) Most of the paper was reworded / restructed to improve
     clarity. Despite the fact that very little of the original text
     is in the revised daft, there are almost no differences in the
     content being presented.

\textbf{I feel the same way about the section on detecting outliers: again, a
well-worked area. Iterative fitting and outlier detection, as an
overall strategy, hav e been used for a long time. I appreciate that
by imposing spatial and temporal conditions you can identify outliers
not otherwise obvi- ous, but it would help to show an example of this,
rather than the current Figure 4, in which most of the identified
cases are ones that any method could easily find.}

### 2017-11-11 ResponseToReviewers/ResponseToReviewers.tex

  8) I expanded the explanation of the REML method, which is used to
     constrain u and eta in Section 3.1 and 3.2.

  9) Most of the paper was reworded / restructed to improve
     clarity. Despite the fact that very little of the original text
     is in the revised daft, there are almost no differences in the
     content being presented.

\textbf{I feel the same way about the section on detecting outliers: again, a
well-worked area. Iterative fitting and outlier detection, as an
overall strategy, hav e been used for a long time. I appreciate that
by imposing spatial and temporal conditions you can identify outliers
not otherwise obvi- ous, but it would help to show an example of this,
rather than the current Figure 4, in which most of the identified
cases are ones that any method could easily find.}

I recognize that outlier detection is a well-worked area. I included
the description of the outlier detection method for completeness, and
not because I thought it was an innovative or superior method. In
order to keep the focus of my paper on my method for computing
transient strain, I move the outlier detection method to the appendix.

In my experimentation with the outlier detection algorithm, it turns
out that the prior spatial covariance has virtually no effect on the
detected outliers. However, the temporal covariance is important. The
prior stochastic model needs to be flexible enough to describe
deformation from, for example, SSEs. This is illustrated in Figure A1
of the revised draft.

@@

\textbf{2+26: possibility not risk: the latter is a technical term in seismic
hazard.}

Done

\textbf{2+47-48: I do not think this description of Shen’s method (which I’ve
used and programmed) is right, since it allows the deformation
gradients to vary in space. I’ve always thought of it as a kind of
adaptive kernel smoother, somewhat like applying 2-D loess; so it is
only depen- dent on nearby points, and has no assumption of uniformity
anywhere. (Actually this raises a more general question: is the method
given here use local or global support?)}

My previous description of Shen's method was poor and misleading. I
have my description in the revised draft.

@@

\textbf{6+56: I suspect that equation (15) took some effort to derive; could
some details go into an Appendix?}

I added a clearer derivation in the revised draft.

\textbf{Figure 6: I like that the strain uncertainties can be shown, but is
there a reason not to plot principle strains using arrows, showing
errors by an ellipse around the tips?}

I have put a lot of thought into representing strain, and I recognize
that strain is conventionally represented by showing the principle
strains axes. However, there is no good way of showing the
uncertainties for the principle strain axes. For example, consider a
perfectly dilational strain that is perturbed by a small amount of
noise. The orientation of principle strains will be highly uncertain,
but the magnitudes of the principle strains will be well constrained.
If we were to draw some sort of confidence interval around the tips of
the principle strains, then they would have a crescent shape (not an
ellipse). It is unclear how I would write a program to draw these
confidence intervals.

Instead of showing the principle strains, I am showing the normal
strains as a function of azimuth. For a given azimuth, the uncertainty
in the normal strain is computed by linear propagation of the
uncertainties in the strain. In my opinion, it is much easier to show
the uncertainties in an azimuthal representation of strain.

I did not include this justification in the revised draft because I
think it would be too much of a digression.

### 2017-11-11 RevisedSubmission/hines_hetland_2017_gji.tex
did more

Crustal strain rates are fundamentally important quantities for
assessing seismic hazard. Knowing where and how quickly strain is
accumulating gives insight into where we can expect stored elastic
energy to be released seismically. Estimates of secular, or long-term,
strain rates have been used to constrain seismic hazard models such as
UCERF3 \citep{Field2014}. Transient crustal strain, which is caused by
geophysical processes such as slow slip events (SSEs) or postseismic
deformation, is also relevant for assessing seismic hazard. While
transient strain itself is not damaging, there is a risk that it can
trigger major earthquakes \citep{Roeloffs2006,Freed2001}.

@@

Displacement data from dense GNSS networks, such as the Plate Boundary
Observatory (PBO), can be used to infer secular and transient strain.
Numerous methods for converting GNSS displacement data to strain have
been proposed, and most involve assuming some parametric form for the
deformation signal. The simplest method for estimating secular strain
rates assumes that GNSS derived velocities can be described with a
first-order polynomial (i.e, the deformation gradients are constant)
over some subnetwork of the GNSS stations \citep[e.g.,][]{Feigl1990,
Murray2000}. The components of the strain rate tensor for each
subnetwork are then determined from the least squares fit to the
observations. The assumption that deformation gradients are spatially
uniform is not appropriate when subnetworks span a very large area. To
help overcome this deficiency, \citet{Shen1996,Shen2015} used an
inverse distance weighting scheme, in which the estimated strain rate
at a point is primarily controlled by observations at nearby stations.
The method of \citet{Shen1996,Shen2015} can be viewed as a form of
local least squares regression with a first-order polynomial. Other
methods for estimating secular strain rates have parameterized GNSS
derived velocities with bi-cubic splines \citep{Beavan2001}, spherical
wavelets \citep{Tape2009}, and elastostatic Green's functions
\citep{Sandwell2016}. The type of basis functions and the number of
degrees of freedom for a parameterization, which are often chosen
subjectively, have a strong influence on the strain solution. If there
are too few degrees of freedom in the parameterization, then estimated
strain rates will be biased and the uncertainties will be
underestimated. On the other hand, if there are too many degrees of
freedom, then there will not be any coherent features in the estimated
strain rates. The methods described by \citet{Beavan2001} and
\citet{Tape2009} also require the user to specify penalty parameters
that control a similar trade-off between bias and variance in the
solution. One could parameterize deformation with a physically
motivated model of interseismic deformation \citep[e.g.,][]{Meade2005,
McCaffrey2007}. In such models the lithospheric rheology and fault
geometries are assumed to be known. Any errors in the assumed physical
model could result in biased strain estimates and underestimated
uncertainties.

@@

The aforementioned studies are concerned with estimating secular
strain rates. In recent years the Southern California Earthquake
Center (SCEC) community has shown interest in developing methods for
detecting transient strain. SCEC supported a transient detection
exercise \citep{Lohman2013}, where several research groups tested
their methods for detecting transient geophysical signal with a
synthetic GNSS dataset. Among the methods tested were the Network
Strain Filter (NSF) \citep{Ohtani2010} and the Network Inversion
Filter (NIF) \citep{Segall1997}. The NSF uses a wavelet
parameterization to describe the spatial component of geophysical
signal. The NIF, which is intended for imaging slow fault slip from
geodetic data, uses the elastic dislocation Green's functions from
\citet{Okada1992}. For the NSF and NIF, the time dependence of the
geophysical signal is modeled as integrated Brownian motion. The
method described in \citet{Holt2013} was also tested in the SCEC
transient detection exercise, which calculates strain rates using a
bi-cubic spatial parameterization of displacements between time
epochs. \citet{Holt2013} defined a detection threshold based on the
strain rate magnitude, and we demonstrate that this is indeed an
effective criterion for identifying geophysical signal. For the same
reasons descibed above, the transient deformation and corresponding
uncertainties estimated by these methods can be biased by the chosen
spatial parameterization. It is then difficult to distinguish signal
from noise with these methods, which limits their utility for
transient detection.

@@

Here we propose using Gaussian process regression (GPR)
\citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR
is closely related to kriging \citep{Cressie1992} and least squares
collocation \citep{Moritz1978}. The latter has been used by
\citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain
rates from GNSS data. GPR is a Bayesian, non-parametric method for
inferring a continuous signal from scattered data. Since GNSS stations
are irregularly spaced and observation times may differ between
stations, GPR is an ideal tool for synthesizing discrete GNSS data
into a spatially and temporally continuous representation of transient
strain. GPR is Bayesian in that we use a prior model to control the
spatial and temporal roughness of the inferred transient strain. The
prior is specified as a stochastic process, namely a Gaussian process.
If there is no information available to help choose an appropriate
prior Gaussian process, then maximum likelihood methods can be used to
objectively choose one that is most consistent with the data. We use
GPR to infer transient strain resulting from SSEs in the Pacific
Northwest, demonstrating that GPR is an effective tool for detecting
transient geophysical processes.

>>>>

Crustal strain rates are fundamentally important quantities for
assessing seismic hazard. Knowing where and how quickly strain is
accumulating gives insight into where we can expect stored elastic
energy to be released seismically. Estimates of secular, or long-term,
strain rates have been used to constrain seismic hazard models such as
UCERF3 \citep{Field2014}. Transient crustal strain, which is caused by
geophysical processes such as slow slip events (SSEs) or postseismic
deformation, is also relevant for assessing seismic hazard. While
transient strain itself is not damaging, there is a possibility that it can
trigger major earthquakes \citep{Roeloffs2006,Freed2001}.

@@

Displacement data from dense GNSS networks, such as the Plate Boundary
Observatory (PBO), can be used to infer secular and transient strain.
Numerous methods for converting GNSS displacement data to strain have
been proposed, and most involve assuming some parametric form for the
deformation signal. The simplest method for estimating secular strain
rates assumes that GNSS derived velocities can be described with a
first-order polynomial (i.e, the deformation gradients are constant)
over some subnetwork of the GNSS stations \citep[e.g.,][]{Feigl1990,
Murray2000}. The components of the strain rate tensor for each
subnetwork are then determined from the least squares fit to the
observations. The assumption that deformation gradients are spatially
uniform is not appropriate when subnetworks span a very large area. To
help overcome this deficiency, \citet{Shen1996,Shen2015} used an
inverse distance weighting scheme, in which the estimated strain rate
at a point is primarily controlled by observations at nearby stations.
The method of \citet{Shen1996,Shen2015} can be viewed as a form of
local least squares regression with a first-order polynomial. Other
methods for estimating secular strain rates have parameterized GNSS
derived velocities with bi-cubic splines \citep{Beavan2001}, spherical
wavelets \citep{Tape2009}, and elastostatic Green's functions
\citep{Sandwell2016}. The type of basis functions and the number of
degrees of freedom for a parameterization, which are often chosen
subjectively, have a strong influence on the strain solution. If there
are too few degrees of freedom in the parameterization, then estimated
strain rates will be biased and the uncertainties will be
underestimated. On the other hand, if there are too many degrees of
freedom, then there will not be any coherent features in the estimated
strain rates. The methods described by \citet{Beavan2001} and
\citet{Tape2009} also require the user to specify penalty parameters
that control a similar trade-off between bias and variance in the
solution. One could also parameterize deformation with a physically
motivated model of interseismic deformation \citep[e.g.,][]{Meade2005,
McCaffrey2007}. In such models the lithospheric rheology and fault
geometries are assumed to be known. Any errors in the assumed physical
model could result in biased strain estimates and underestimated
uncertainties.

@@

The aforementioned studies are concerned with estimating secular
strain rates. In recent years the Southern California Earthquake
Center (SCEC) community has shown interest in developing methods for
detecting transient strain. SCEC supported a transient detection
exercise \citep{Lohman2013}, where several research groups tested
their methods for detecting transient geophysical signal with a
synthetic GNSS dataset. Among the methods tested were the Network
Strain Filter (NSF) \citep{Ohtani2010} and the Network Inversion
Filter (NIF) \citep{Segall1997}. The NSF uses a wavelet
parameterization to describe the spatial component of geophysical
signal. The NIF, which is intended for imaging slow fault slip from
geodetic data, uses the elastic dislocation Green's functions from
\citet{Okada1992}. For the NSF and NIF, the time dependence of the
geophysical signal is modeled as integrated Brownian motion. The
method described in \citet{Holt2013} was also tested in the SCEC
transient detection exercise, which calculates strain rates using a
bi-cubic spatial parameterization of displacements between time
epochs. \citet{Holt2013} defined a detection threshold based on the
strain rate magnitude, and we demonstrate that this is indeed an
effective criterion for identifying a geophysical signal. For the same
reasons descibed above, the transient deformation and corresponding
uncertainties estimated by these methods can be biased by the chosen
spatial parameterization. It is then difficult to distinguish signal
from noise with these methods, which limits their utility for
transient detection.

@@

Here we propose using Gaussian process regression (GPR)
\citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR
is closely related to kriging \citep{Cressie1992} and least squares
collocation \citep{Moritz1978}. The latter has been used by
\citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain
rates from GNSS data. GPR is a Bayesian, non-parametric method for
inferring a continuous function from scattered data. Since GNSS
stations are irregularly spaced and observation times may differ
between stations, GPR is an ideal tool for synthesizing discrete GNSS
data into a spatially and temporally continuous representation of
transient strain. GPR is Bayesian in that we use a prior model to
control the spatial and temporal roughness of the inferred transient
strain. The prior is specified as a stochastic process, namely a
Gaussian process. If there is no information available to help choose
an appropriate prior Gaussian process, then maximum likelihood methods
can be used to objectively choose one that is most consistent with the
data. We use GPR to infer transient strain resulting from SSEs in the
Pacific Northwest, demonstrating that GPR is an effective tool for
detecting transient geophysical processes.
